The Talking Heads Experiment

### 2015-04-10 chap5/chap5.tex
cumulative changes to preface and chapter 1
updated labels, image widths, quotation marks, capitalisation, etc. for chapters 1-7

%\documentclass{../_combined/fcg-book}
\chapter{The Naming Game}

@@

Immediately we observe a major difficulty. In realistic
language games, where agents cannot inspect each others' 
brain states nor transmit meanings directly, 
there is no feedback about the 
meaning of a word, only about the referent. For example, 
when a speaker says `wabo' and the hearer has correctly pointed 
to the referent that the speaker intended, neither the 
speaker nor the hearer can know whether they were 
using the same meaning. They can only know that they 
arrived at the same referent. When a game fails, the speaker 
can only point to the topic and the hearer then 
tries to figure out what possible meaning 
could have been applicable. The speaker cannot
communicate directly the `right' meaning, and very often more 
than one meaning is possible to distinguish a topic from 
other objects in the context, so that the hearer will 
not necessarily guess the meaning used by the speaker. 
I will call this the gavagai-problem, because Quine 
used this word to illustrate exactly this difficulty.\footnote{See \cite{Quine:1960}, p. 29,30.}
Quine evoked the problem of an anthropologist
trying to figure out what a native speaking an unknown
language might mean when he utters `gavagai' while pointing
to a white rabbit scurrying by.

@@

\begin{table}
\begin{center}
\begin{tabular}{ l  l }
\lsptoprule
{\itshape meaning} & {\itshape form} \\ 
\midrule{}
{}[DARK] & coba \\ 
{}[LARGE] & wabidu \\ 
... & ... \\ 
\lspbottomrule
\end{tabular}
\end{center}
\caption{\label{tab:t-mem} Associative memory of a single agent.}
\end{table}
As an agent is acquiring his lexicon, there are going to 
be stages when he is not yet sure about the meaning of 
a certain form. So it must be possible for the agent
to store different 
meanings for the same form and different forms for the
same meaning. This can easily be done by extending the memory capacity
to cross-associate multiple items. Agents can then 
handle amiguity (one word can have different
meanings) and synonymy (one meaning can 
be associated with many different words).

@@

\begin{table}
\begin{center}
\begin{tabular}{c c c c  c  c  c } \lsptoprule 
& coba & zapo & bila & pama & wabidu & limiri \\ 
\midrule {}		
{}[DARK] & 0.3 & 0.2 & 0.1 & 0.8 & - & - \\ 
{}[LARGE] & - & - & - & 0.5 & 0.3 & 0.6 \\ 
\lspbottomrule
\end{tabular}
\caption{\label{tab:mem2} Example lexicon with multiple associations.}
\end{center}
\end{table}
From this table we see that the agent prefers to use the 
word `pama' for [DARK] and `limiri' for [LARGE].

@@

Every time an agent successfully uses a form-meaning pair 
for speaking, he increments
the score with a specific increment $\delta$. 
$\delta$ is relatively high, typically equal to 0.1. 
The scores of competing associations, i.e. 
associations that used another form for the same meaning
are decremented with $\delta$. The score remains however bounded 
between 0.0 and 1.0. This way the `best' 
form-meaning pair stands out more clearly next time 
around.\footnote{A systematic investigation of alternatives for 
the updating function is contained in \cite{Oliphant:1997}. 
The dynamics of the mechanisms used in the 
Talking Heads experiments are being investigated in the
Ph.D thesis of Frederic Kaplan.}

@@

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.60\textwidth]{chap5/figs/incr-decr}}
\caption{ \label{incr-decr} 
Score adjustments after a successful game. Scores of used 
associations go up and their competitors go down.}
\end{figure}

@@

When virgin agents start playing naming games, their 
associative memories are completely empty. Each agent 
needs two additional activities to get a lexicon
emerging: 
\begin{itemize}
\item 
When an agent does not have a word for a
meaning he wants to communicate, he is allowed to create
a new word (by random combination of vowels and 
consonants) and add that to his 
lexicon. Agents are assumed to have a 
shared repertoire of syllables which they can all 
produce and recognise. This happens with a certain 
probability, the {\itshape word creation rate} $w_{c}$. 
This rate reflects how `free' the agent feels to 
extend the lexicon. 
\item 
When an agent hears a word he has 
never heard before, he may add this new word to his 
repertoire. Again this happens with a certain
probability, the {\itshape word absorption rate} $w_{a}$. 
This rate reflects the critical attitude with which an 
agent accepts the linguistic authority of other agents. 
\end{itemize}
These rates are not critical but must of course
be positive. Experiments continue to work 
when agents always make a new word ($w_{c}=1$) 
and always absorb the word of the other ($w_{a}=1$).

@@

\begin{table}
\begin{center}
\begin{tabular}{l  l  l }
\lsptoprule 
{\itshape Meaning} & {\itshape Word} & {\itshape Score} \\ \midrule
{}[LARGE]& pida& 0.20\\ 
  & fobu& 0.1\\ 
{}[LIGHT]& gi& 0.0\\ 
{}[SMALL]& di & 0.10\\ 
\lspbottomrule
\end{tabular}
\caption{\label{tab:t-mem3} Associative memory of a single agent after 100 games.}
\end{center}
\end{table}
Both the words `di' and `pida' are present but with 
weak scores. There are two synonyms for 
{}[LARGE]: `pida' and `fobu', but `pida' is 
preferred. There is a word `gi' who is available 
for [LIGHT] but does not have a positive score,
because successive trials failed to yield a successful game.

A table such as the one above only represents the lexicon
of a single agent. It is highly unlikely that two agents
share the same lexicon because each agent will have 
had different encounters and hence different language
experiences. A picture of the lexicon of the group from the 
viewpoint of an outside observer can be obtained by
inspecting the internal states of each agent to 
construct the {\itshape group lexicon}. It groups the 
dominating meaning-form associations for all possible
meanings and the frequency of each association. 
It gives a picture of `the' lexicon 
in the group. The group lexicon
for the complete population of five agents after 50 language games
is shown in \tabref{tab:t-mem4}.

\begin{table}
\begin{center}
\begin{tabular}{l  l  l }
\lsptoprule 
{\itshape Meaning}& {\itshape Word} & {\itshape Frequency} \\ \midrule 
{}[LARGE]& pida& 0.40\\ 
{}[LIGHT]& gi & 0.60\\ 
{}[SMALL]& di & 0.40\\ 
\lspbottomrule
\end{tabular}
\caption{\label{tab:t-mem4} Population lexicon after 50 games}
\end{center}
\end{table}
This reflects a situation where 40 \%
of the agents prefers to name [LARGE] with
`pida'. 60 \% of the agents
use `gi' for [LIGHT], and 40 \% `di' for [SMALL]. 
The other meanings ([DARK] and [RED]) do not have names yet.

@@

\begin{table}
\begin{center}
\begin{tabular}{l  l  c  l  l  c } \midrule 
{\itshape meaning} & {\itshape form} & {\itshape frequency} & {\itshape meaning} & {\itshape form} & {\itshape frequency}\\  
{}[DARK]& go & 1.00 & [LARGE]& pida & 1.00 \\  
{}[LIGHT]& gi & 1.00 & [SMALL]& di & 1.00 \\  
{}[RED]& ga & 1.00 & - & - & -  \\  
\lspbottomrule
\end{tabular}
\caption{\label{tab:t-mem5} Consensus in group lexicon, reached after 250 games.}
\end{center}
\end{table}
Once the agents have reached this stage, the lexicon does not 
change anymore, because all agents now prefer the same word for 
each possible meaning and would never choose another one
nor encounter another one.

@@

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.70\textwidth]{chap5/figs/success}}
\caption{ \label{success} 
The graph displays on the y-axis the average success every 10 games
in a population of five agents lexicalising five
different meanings. The x-axis 
shows the number of games. Average success
rapidly climbs until it reaches total success after 
about 180 games.}
\end{figure}

@@

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.70\textwidth]{chap5/figs/larger}}
\caption{\footnotesize \label{larger} 
The graph displays the evolution of communicative success
for larger and larger populations. The
number of games on the x-axis
is divided by the number of agents. }
\end{figure}
We see for example that for 20 agents and 20 meanings 
success climbs to total success after about 10,000 games.
This is still surprisingly low particularly as
success is
are already above 95 \% after about 5000 games. Games can be played
in parallel by different agents because the system 
is entirely distributed. If we divide
the number of agents by the number of games, we see that about 250
games are needed by the agents to get 95 \% success, which means
that every meaning needs to appear about 10 times for each agent. 
Interestingly enough, the larger the population of agents, 
the more the success curve approximates an 
S-shape, which has 
been observed empirically in the spreading of new
linguistic conventions. The same curve shape is familiar to
biologists studying models of competitive growth, suggesting 
a strong relationships between ecological dynamics and language\is{S-shape curve}
spreading.\footnote{The S-shaped curve is discussed in \cite{McMahon:1994} p. 52. For examples
of biological models with similar properties, see \cite{May:1976}.}

@@

\begin{table}
\begin{center}
\begin{tabular}{l  l  l  l  l  l }
\lsptoprule 
{\itshape meaning} & {\itshape form} & {\itshape frequency} & {\itshape meaning} & {\itshape form} & {\itshape frequency} \\ \midrule
{}[DARK]& dato & 1.00 & [LARGE]& biti & 0.80 \\  
{}[LIGHT]& pitu & 0.60 & [SMALL]& dopu & 1.00 \\  
{}[RED]& gabi & 1.00 & [GREEN]& gu & 0.85 \\  
{}[SQUARE]& koti & 0.50 & [RECTANGLE]& totu & 0.65 \\  
{}[LEFT]& toga & 0.90 & [BLUE]& ku & 0.80 \\  
{}[YELLOW]& gubo & 0.55 & [CHARMING]& ge & 1.00 \\  
{}[TRIANGLE]& bu & 0.85 & [SQUARE]& ba & 0.60 \\  
{}[FAST]& beke & 1.00 & [SLOW]& tu & 0.95 \\  
{}[CIRCLE]& ke & 0.75 & [RIGHT]& gaba & 0.95 \\  
{}[UP]& butu & 1.00 & [DOWN]& ki & 0.95 \\  
\lspbottomrule
\end{tabular}
\caption{\label{tab:t-mem3000} Group lexicon after 3000 games.}
\end{center}
\end{table}
The lexical coherence is at this point equal to 0.835.

@@

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.70\textwidth]{chap5/figs/suc-coh}}
\caption{\label{suc-coh}This figure shows the evolution of both average
success and lexical coherence for a group of 20 agents and 20 
meanings. Total lexical coherence climbs more slowly
once the 
population has reached a high average success.}
\end{figure}
Does total success imply that all agents use the 
same lexicon? Not really. To have success, the hearer
must associate the form used by the speaker
with the same meaning. But it is not required that the
hearer himself prefers to use the same form for the same meaning, 
synonyms may occur.  
A speaker of British English typically uses the word `pavement', 
whereas an American prefers `sidewalk', even though he
understands `pavement'. Thus there can 
be several forms active in the same population, even though 
the outcome of a game is always successful. We will see 
later that synonyms do get damped, as is the case in 
human natural lexicons.

@@

\begin{table}
\begin{center}
\begin{tabular}{l  l  c  l  l  c } \lsptoprule 
{\itshape meaning} & {\itshape form} & {\itshape frequency} & {\itshape meaning} & {\itshape form} & {\itshape frequency}\\ \midrule 
{}[M1]& gebo & 1.00 & [M2]& goge & 1.00 \\  
{}[M3]& koto & 0.70 & [M4]& da & 1.00 \\  
{}[M5]& peko & 1.00 & [M6]& ki & 1.00 \\  
{}[M7]& gipe & 1.00 & [M8]& kedo & 1.00 \\  
{}[M9]& do & 1.00 & [M10]& gige & 1.00 \\  
{}[M11]& pi & 1.00 & [M12]& bu & 1.00 \\  
{}[M13]& pa & 1.00 & [M14]& kipa & 1.00 \\  
{}[M15]& depi & 0.95 & [M16]& pudi & 1.00 \\  
{}[M17]& tegi & 1.00 & [M18]& ba & 0.90 \\  
{}[M19]& ko & 1.00 & [M20]& guda & 1.00 \\  
 \lspbottomrule
\end{tabular}
\caption{\label{tab:gphase1} Group lexicon after first phase.}
\end{center}
\end{table}
In phase 2, a relatively small meaning flux is introduced (one
new meaning every 1000 games). As can 
be seen from \figref{inoutmean}, 
the population copes with the change. New words are created and 
propagate in the population. The following group lexicon
shows that for newcomers like [M22] or [M25] a total 
consensus has emerged. Words for the latest
new meanings, [M28] and [M29], still have low frequencies.

\begin{table}
\begin{center}
\begin{tabular}{ l  l  c  l  l  c } \lsptoprule 
{\itshape meaning} & {\itshape form} & {\itshape frequency} & {\itshape meaning} & {\itshape form} & {\itshape frequency}\\ \midrule 
{}[M1]& gebo & 1.00 & [M2]& goge & 1.00 \\  
{}[M3]& koto & 1.00 & [M4]& da & 1.00 \\  
{}[M5]& peko & 1.00 & [M6]& ki & 1.00 \\  
{}[M7]& gipe & 1.00 & [M8]& kedo & 1.00 \\  
{}[M9]& do & 1.00 & [M10]& gige & 1.00 \\  
{}[M11]& pi & 1.00 & [M12]& bu & 1.00 \\  
{}[M13]& pa & 1.00 & [M14]& kipa & 1.00 \\  
{}[M15]& depi & 1.00 & [M16]& pudi & 1.00 \\  
{}[M17]& tegi & 1.00 & [M18]& ba & 1.00 \\  
{}[M19]& ko & 1.00 & [M20]& guda & 1.00 \\  
{}[M22]& to & 1.00 & [M23]& de & 0.85 \\  
{}[M24]& tabo & 0.95 & [M25]& piku & 1.00 \\  
{}[M26]& ku & 1.00 & [M27]& pugu & 1.00 \\  
{}[M28]& tete & 0.35 & [M29]& todu & 0.35 \\  
\lspbottomrule
\end{tabular}
\caption{\label{tab:phase1} Group lexicon after second phase.}
\end{center}
\end{table}
Next (phase 3 in \figref{inoutmean}) a much higher 
meaning flux is imposed (one new meaning every 100 games). 
Lexical coherence decreases and
average success plummets. There is not enough time to propagate
the new conventions in the group. 
Note that lexical coherence drops slower than 
success when the lexicon disintegrates. 
Coherence is based on the average for all 
meanings, thus only the new ones are therefore affecting overall coherence. 
Success drops more rapidly 
because of the high rate of failure of new meanings.

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.70\textwidth]{chap5/figs/in+out}}
\caption{ \label{inoutmean}  Both average success (every 100 games) and lexical coherence 
is shown in cases of an inflow of meanings for a population 
of 20 agents starting with 20 meanings (phase 1). The inflow 
is 1/1000 in phase 2, 1/100 in phase 3 and 1/1000 in phase 4.}
\end{figure}

@@

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.70\textwidth]{chap5/figs/birth}}
\caption{\label{birth} Evolution of communicative 
success with different birth-rates, starting from a
population of 20 agents (phase 1). 
Next the birth rate is increased from 1 new agent 
every 1000 games (phase 2) to 
1 new agent every 100 games (phase 3), 
and then to 1 every 50 games (phase 4).}
\end{figure}

@@

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.70\textwidth]{chap5/figs/birth+death}}
\caption{\footnotesize \label{birth+death} Success and population size 
is shown for a series of 35,000 language games. 
The population starts with 20 agents and 20 
meanings (phase 1). Then an influx and outflux is introduced 
at the rate of 1/1000 games (phase 2). The lexicon
maintains itself. In phase 3 agents enter and leave at
the rate of 1/100 games. Success lowers. In phase 
4 the rate of change is brought back to 1/1000 games
and success is regained.}
\end{figure}

@@

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.70\textwidth]{chap5/figs/comp1}}
\caption{\label{competition} Simulation with 
a population of 20 agents. The meaning-form competition 
diagram shows the frequency of all competing forms for a 
single meaning. We see a winner-take-all situation
with one word (`pe') dominating.}
\end{figure}

@@

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.40
\textwidth]{chap5/figs/fig-agent-distribution}}
\caption{\footnotesize The figure shows the spatial distribution of a set
of 20 agents. There is clustering around three centers.}
\label{figure-agent-distribution}
\end{figure}

@@

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.65\textwidth]{chap5/figs/comm-succ}}
\caption{\footnotesize Evolution of average communicative success per 25 games
in a group of agents with first weak (phase 1) and then strong
interactions (phase 2).}
\label{figure-communicative-success-in-space}
\end{figure}

@@

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.80\textwidth]{chap5/figs/coherence}}
\caption{Evolution of coherence in the total population 
and in the individual clusters is shown on the same
scale as the previous figure. When contact 
is increased (phase 2), global coherence begins to rise steadily.}
\label{figure-coherence-in-space}
\end{figure}
Simulations show that, just as in human languages,
increased contact causes at first a rapid increase in
bilingualism, then a gradual mixing of the languages, and, 
if the contact
continues, an evolution towards complete coherence. 
The more rapid the contact is
increased, the faster the three phases can be observed.

>>>>

%\documentclass{../_combined/fcg-book}
\chapter{The Naming Game}
\label{chap:5}

@@

Immediately we observe a major difficulty. In realistic
language games, where agents cannot inspect each others' 
brain states nor transmit meanings directly, 
there is no feedback about the 
meaning of a word, only about the referent. For example, 
when a speaker says `wabo' and the hearer has correctly pointed 
to the referent that the speaker intended, neither the 
speaker nor the hearer can know whether they were 
using the same meaning. They can only know that they 
arrived at the same referent. When a game fails, the speaker 
can only point to the topic and the hearer then 
tries to figure out what possible meaning 
could have been applicable. The speaker cannot
communicate directly the ``right'' meaning, and very often more 
than one meaning is possible to distinguish a topic from 
other objects in the context, so that the hearer will 
not necessarily guess the meaning used by the speaker. 
I will call this the gavagai-problem, because Quine 
used this word to illustrate exactly this difficulty.\footnote{See \cite{Quine:1960}, p. 29,30.}
Quine evoked the problem of an anthropologist
trying to figure out what a native speaking an unknown
language might mean when he utters `gavagai' while pointing
to a white rabbit scurrying by.

@@

\begin{table}
\begin{center}
\begin{tabular}{ l  l }
\lsptoprule
{\itshape meaning} & {\itshape form} \\ 
\midrule{}
{}[DARK] & coba \\ 
{}[LARGE] & wabidu \\ 
... & ... \\ 
\lspbottomrule
\end{tabular}
\end{center}
\caption{\label{tab:t-mem}Associative memory of a single agent.}
\end{table}
As an agent is acquiring his lexicon, there are going to 
be stages when he is not yet sure about the meaning of 
a certain form. So it must be possible for the agent
to store different 
meanings for the same form and different forms for the
same meaning. This can easily be done by extending the memory capacity
to cross-associate multiple items. Agents can then 
handle amiguity (one word can have different
meanings) and synonymy (one meaning can 
be associated with many different words).

@@

\begin{table}
\begin{center}
\begin{tabular}{c c c c  c  c  c } \lsptoprule 
& coba & zapo & bila & pama & wabidu & limiri \\ 
\midrule {}		
{}[DARK] & 0.3 & 0.2 & 0.1 & 0.8 & - & - \\ 
{}[LARGE] & - & - & - & 0.5 & 0.3 & 0.6 \\ 
\lspbottomrule
\end{tabular}
\caption{\label{tab:mem2}Example lexicon with multiple associations.}
\end{center}
\end{table}
From this table we see that the agent prefers to use the 
word `pama' for [DARK] and `limiri' for [LARGE].

@@

Every time an agent successfully uses a form-meaning pair 
for speaking, he increments
the score with a specific increment $\delta$. 
$\delta$ is relatively high, typically equal to 0.1. 
The scores of competing associations, i.e. 
associations that used another form for the same meaning
are decremented with $\delta$. The score remains however bounded 
between 0.0 and 1.0. This way the ``best'' 
form-meaning pair stands out more clearly next time 
around.\footnote{A systematic investigation of alternatives for 
the updating function is contained in \cite{Oliphant:1997}. 
The dynamics of the mechanisms used in the 
Talking Heads experiments are being investigated in the
Ph.D thesis of Frederic Kaplan.}

@@

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.60\textwidth]{chap5/figs/incr-decr.pdf}}
\caption{\label{incr-decr} 
Score adjustments after a successful game. Scores of used 
associations go up and their competitors go down.}
\end{figure}

@@

When virgin agents start playing naming games, their 
associative memories are completely empty. Each agent 
needs two additional activities to get a lexicon
emerging: 
\begin{itemize}
\item 
When an agent does not have a word for a
meaning he wants to communicate, he is allowed to create
a new word (by random combination of vowels and 
consonants) and add that to his 
lexicon. Agents are assumed to have a 
shared repertoire of syllables which they can all 
produce and recognise. This happens with a certain 
probability, the {\itshape word creation rate} $w_{c}$. 
This rate reflects how ``free'' the agent feels to 
extend the lexicon. 
\item 
When an agent hears a word he has 
never heard before, he may add this new word to his 
repertoire. Again this happens with a certain
probability, the {\itshape word absorption rate} $w_{a}$. 
This rate reflects the critical attitude with which an 
agent accepts the linguistic authority of other agents. 
\end{itemize}
These rates are not critical but must of course
be positive. Experiments continue to work 
when agents always make a new word ($w_{c}=1$) 
and always absorb the word of the other ($w_{a}=1$).

@@

\begin{table}
\begin{center}
\begin{tabular}{l  l  l }
\lsptoprule 
{\itshape Meaning} & {\itshape Word} & {\itshape Score} \\ \midrule
{}[LARGE]& pida& 0.20\\ 
  & fobu& 0.1\\ 
{}[LIGHT]& gi& 0.0\\ 
{}[SMALL]& di & 0.10\\ 
\lspbottomrule
\end{tabular}
\caption{\label{tab:t-mem3}Associative memory of a single agent after 100 games.}
\end{center}
\end{table}
Both the words `di' and `pida' are present but with 
weak scores. There are two synonyms for 
{}[LARGE]: `pida' and `fobu', but `pida' is 
preferred. There is a word `gi' who is available 
for [LIGHT] but does not have a positive score,
because successive trials failed to yield a successful game.

A table such as the one above only represents the lexicon
of a single agent. It is highly unlikely that two agents
share the same lexicon because each agent will have 
had different encounters and hence different language
experiences. A picture of the lexicon of the group from the 
viewpoint of an outside observer can be obtained by
inspecting the internal states of each agent to 
construct the {\itshape group lexicon}. It groups the 
dominating meaning-form associations for all possible
meanings and the frequency of each association. 
It gives a picture of ``the'' lexicon 
in the group. The group lexicon
for the complete population of five agents after 50 language games
is shown in \tabref{tab:t-mem4}.

\begin{table}
\begin{center}
\begin{tabular}{l  l  l }
\lsptoprule 
{\itshape Meaning}& {\itshape Word} & {\itshape Frequency} \\ \midrule 
{}[LARGE]& pida& 0.40\\ 
{}[LIGHT]& gi & 0.60\\ 
{}[SMALL]& di & 0.40\\ 
\lspbottomrule
\end{tabular}
\caption{\label{tab:t-mem4}Population lexicon after 50 games}
\end{center}
\end{table}
This reflects a situation where 40 \%
of the agents prefers to name [LARGE] with
`pida'. 60 \% of the agents
use `gi' for [LIGHT], and 40 \% `di' for [SMALL]. 
The other meanings ([DARK] and [RED]) do not have names yet.

@@

\begin{table}
\begin{center}
\begin{tabular}{l  l  c  l  l  c } \midrule 
{\itshape meaning} & {\itshape form} & {\itshape frequency} & {\itshape meaning} & {\itshape form} & {\itshape frequency}\\  
{}[DARK]& go & 1.00 & [LARGE]& pida & 1.00 \\  
{}[LIGHT]& gi & 1.00 & [SMALL]& di & 1.00 \\  
{}[RED]& ga & 1.00 & - & - & -  \\  
\lspbottomrule
\end{tabular}
\caption{\label{tab:t-mem5}Consensus in group lexicon, reached after 250 games.}
\end{center}
\end{table}
Once the agents have reached this stage, the lexicon does not 
change anymore, because all agents now prefer the same word for 
each possible meaning and would never choose another one
nor encounter another one.

@@

\begin{figure}[htbp]
  \centerline{\includegraphics[width=\textwidth]{chap5/figs/success.pdf}}
\caption{\label{success} 
The graph displays on the y-axis the average success every 10 games
in a population of five agents lexicalising five
different meanings. The x-axis 
shows the number of games. Average success
rapidly climbs until it reaches total success after 
about 180 games.}
\end{figure}

@@

\begin{figure}[htbp]
  \centerline{\includegraphics[width=\textwidth]{chap5/figs/larger.pdf}}
\caption{\label{larger} 
The graph displays the evolution of communicative success
for larger and larger populations. The
number of games on the x-axis
is divided by the number of agents. }
\end{figure}
We see for example that for 20 agents and 20 meanings 
success climbs to total success after about 10,000 games.
This is still surprisingly low particularly as
success is
are already above 95 \% after about 5000 games. Games can be played
in parallel by different agents because the system 
is entirely distributed. If we divide
the number of agents by the number of games, we see that about 250
games are needed by the agents to get 95 \% success, which means
that every meaning needs to appear about 10 times for each agent. 
Interestingly enough, the larger the population of agents, 
the more the success curve approximates an 
S-shape, which has 
been observed empirically in the spreading of new
linguistic conventions. The same curve shape is familiar to
biologists studying models of competitive growth, suggesting 
a strong relationships between ecological dynamics and language\is{S-shape curve}
spreading.\footnote{The S-shaped curve is discussed in \cite{McMahon:1994} p. 52. For examples
of biological models with similar properties, see \cite{May:1976}.}

@@

\begin{table}
\begin{center}
\begin{tabular}{l  l  l  l  l  l }
\lsptoprule 
{\itshape meaning} & {\itshape form} & {\itshape frequency} & {\itshape meaning} & {\itshape form} & {\itshape frequency} \\ \midrule
{}[DARK]& dato & 1.00 & [LARGE]& biti & 0.80 \\  
{}[LIGHT]& pitu & 0.60 & [SMALL]& dopu & 1.00 \\  
{}[RED]& gabi & 1.00 & [GREEN]& gu & 0.85 \\  
{}[SQUARE]& koti & 0.50 & [RECTANGLE]& totu & 0.65 \\  
{}[LEFT]& toga & 0.90 & [BLUE]& ku & 0.80 \\  
{}[YELLOW]& gubo & 0.55 & [CHARMING]& ge & 1.00 \\  
{}[TRIANGLE]& bu & 0.85 & [SQUARE]& ba & 0.60 \\  
{}[FAST]& beke & 1.00 & [SLOW]& tu & 0.95 \\  
{}[CIRCLE]& ke & 0.75 & [RIGHT]& gaba & 0.95 \\  
{}[UP]& butu & 1.00 & [DOWN]& ki & 0.95 \\  
\lspbottomrule
\end{tabular}
\caption{\label{tab:t-mem3000}Group lexicon after 3000 games.}
\end{center}
\end{table}
The lexical coherence is at this point equal to 0.835.

@@

\begin{figure}[htbp]
  \centerline{\includegraphics[width=\textwidth]{chap5/figs/suc-coh.pdf}}
\caption{\label{suc-coh}This figure shows the evolution of both average
success and lexical coherence for a group of 20 agents and 20 
meanings. Total lexical coherence climbs more slowly
once the 
population has reached a high average success.}
\end{figure}
Does total success imply that all agents use the 
same lexicon? Not really. To have success, the hearer
must associate the form used by the speaker
with the same meaning. But it is not required that the
hearer himself prefers to use the same form for the same meaning, 
synonyms may occur.  
A speaker of British English typically uses the word `pavement', 
whereas an American prefers `sidewalk', even though he
understands `pavement'. Thus there can 
be several forms active in the same population, even though 
the outcome of a game is always successful. We will see 
later that synonyms do get damped, as is the case in 
human natural lexicons.

@@

\begin{table}
\begin{center}
\begin{tabular}{l  l  c  l  l  c } \lsptoprule 
{\itshape meaning} & {\itshape form} & {\itshape frequency} & {\itshape meaning} & {\itshape form} & {\itshape frequency}\\ \midrule 
{}[M1]& gebo & 1.00 & [M2]& goge & 1.00 \\  
{}[M3]& koto & 0.70 & [M4]& da & 1.00 \\  
{}[M5]& peko & 1.00 & [M6]& ki & 1.00 \\  
{}[M7]& gipe & 1.00 & [M8]& kedo & 1.00 \\  
{}[M9]& do & 1.00 & [M10]& gige & 1.00 \\  
{}[M11]& pi & 1.00 & [M12]& bu & 1.00 \\  
{}[M13]& pa & 1.00 & [M14]& kipa & 1.00 \\  
{}[M15]& depi & 0.95 & [M16]& pudi & 1.00 \\  
{}[M17]& tegi & 1.00 & [M18]& ba & 0.90 \\  
{}[M19]& ko & 1.00 & [M20]& guda & 1.00 \\  
 \lspbottomrule
\end{tabular}
\caption{\label{tab:gphase1}Group lexicon after first phase.}
\end{center}
\end{table}
In phase 2, a relatively small meaning flux is introduced (one
new meaning every 1000 games). As can 
be seen from \figref{inoutmean}, 
the population copes with the change. New words are created and 
propagate in the population. The following group lexicon
shows that for newcomers like [M22] or [M25] a total 
consensus has emerged. Words for the latest
new meanings, [M28] and [M29], still have low frequencies.

\begin{table}
\begin{center}
\begin{tabular}{ l  l  c  l  l  c } \lsptoprule 
{\itshape meaning} & {\itshape form} & {\itshape frequency} & {\itshape meaning} & {\itshape form} & {\itshape frequency}\\ \midrule 
{}[M1]& gebo & 1.00 & [M2]& goge & 1.00 \\  
{}[M3]& koto & 1.00 & [M4]& da & 1.00 \\  
{}[M5]& peko & 1.00 & [M6]& ki & 1.00 \\  
{}[M7]& gipe & 1.00 & [M8]& kedo & 1.00 \\  
{}[M9]& do & 1.00 & [M10]& gige & 1.00 \\  
{}[M11]& pi & 1.00 & [M12]& bu & 1.00 \\  
{}[M13]& pa & 1.00 & [M14]& kipa & 1.00 \\  
{}[M15]& depi & 1.00 & [M16]& pudi & 1.00 \\  
{}[M17]& tegi & 1.00 & [M18]& ba & 1.00 \\  
{}[M19]& ko & 1.00 & [M20]& guda & 1.00 \\  
{}[M22]& to & 1.00 & [M23]& de & 0.85 \\  
{}[M24]& tabo & 0.95 & [M25]& piku & 1.00 \\  
{}[M26]& ku & 1.00 & [M27]& pugu & 1.00 \\  
{}[M28]& tete & 0.35 & [M29]& todu & 0.35 \\  
\lspbottomrule
\end{tabular}
\caption{\label{tab:phase1}Group lexicon after second phase.}
\end{center}
\end{table}
Next (phase 3 in \figref{inoutmean}) a much higher 
meaning flux is imposed (one new meaning every 100 games). 
Lexical coherence decreases and
average success plummets. There is not enough time to propagate
the new conventions in the group. 
Note that lexical coherence drops slower than 
success when the lexicon disintegrates. 
Coherence is based on the average for all 
meanings, thus only the new ones are therefore affecting overall coherence. 
Success drops more rapidly 
because of the high rate of failure of new meanings.

\begin{figure}[htbp]
  \centerline{\includegraphics[width=\textwidth]{chap5/figs/in+out.pdf}}
\caption{\label{inoutmean}Both average success (every 100 games) and lexical coherence 
is shown in cases of an inflow of meanings for a population 
of 20 agents starting with 20 meanings (phase 1). The inflow 
is 1/1000 in phase 2, 1/100 in phase 3 and 1/1000 in phase 4.}
\end{figure}

@@

\begin{figure}[htbp]
  \centerline{\includegraphics[width=\textwidth]{chap5/figs/birth.pdf}}
\caption{\label{birth}Evolution of communicative 
success with different birth-rates, starting from a
population of 20 agents (phase 1). 
Next the birth rate is increased from 1 new agent 
every 1000 games (phase 2) to 
1 new agent every 100 games (phase 3), 
and then to 1 every 50 games (phase 4).}
\end{figure}

@@

\begin{figure}[htbp]
  \centerline{\includegraphics[width=\textwidth]{chap5/figs/birth+death.pdf}}
\caption{\label{birth+death}Success and population size 
is shown for a series of 35,000 language games. 
The population starts with 20 agents and 20 
meanings (phase 1). Then an influx and outflux is introduced 
at the rate of 1/1000 games (phase 2). The lexicon
maintains itself. In phase 3 agents enter and leave at
the rate of 1/100 games. Success lowers. In phase 
4 the rate of change is brought back to 1/1000 games
and success is regained.}
\end{figure}

@@

\begin{figure}[htbp]
  \centerline{\includegraphics[width=\textwidth]{chap5/figs/comp1.pdf}}
\caption{\label{competition}Simulation with 
a population of 20 agents. The meaning-form competition 
diagram shows the frequency of all competing forms for a 
single meaning. We see a winner-take-all situation
with one word (`pe') dominating.}
\end{figure}

@@

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.40
\textwidth]{chap5/figs/fig-agent-distribution.pdf}}
\caption{The figure shows the spatial distribution of a set
of 20 agents. There is clustering around three centers.}
\label{figure-agent-distribution}
\end{figure}

@@

\begin{figure}[htbp]
  \centerline{\includegraphics[width=\textwidth]{chap5/figs/comm-succ.pdf}}
\caption{Evolution of average communicative success per 25 games
in a group of agents with first weak (phase 1) and then strong
interactions (phase 2).}
\label{figure-communicative-success-in-space}
\end{figure}

@@

\begin{figure}[htbp]
  \centerline{\includegraphics[width=\textwidth]{chap5/figs/coherence.pdf}}
\caption{Evolution of coherence in the total population 
and in the individual clusters is shown on the same
scale as the previous figure. When contact 
is increased (phase 2), global coherence begins to rise steadily.}
\label{figure-coherence-in-space}
\end{figure}
Simulations show that, just as in human languages,
increased contact causes at first a rapid increase in
bilingualism, then a gradual mixing of the languages, and, 
if the contact
continues, an evolution towards complete coherence. 
The more rapid the contact is
increased, the faster the three phases can be observed.
